---
title: "Solid Tumor Example"
description: |
  A workflow to develop a classifier of a subset of solid tumors 
  from methylation array data.
author:
  - name: Mark Chen
    url: https://markgene.github.io/
    affiliation: SJCRH
    affiliation_url: https://www.stjude.org/
date: "`r Sys.Date()`"
citation_url: https://github.com/markgene/yamatClassifier
output: 
  radix::radix_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

```{r load_lib, message=FALSE, echo=FALSE, eval=TRUE}
library(yamat)
library(yamatClassifier)
library(ggplot2)
library(ggpubr)
library(corrplot)
library(RColorBrewer)
library(plyr)
library(dplyr)
library(RSpectra)
```


# Data

I use the data from the paper by S. Peter Wu et al. entitled [DNA Methylation–Based Classifier for Accurate Molecular Diagnosis of Bone Sarcomas](http://ascopubs.org/doi/full/10.1200/PO.17.00031). It has 36 samples. However, I borrow the methodology manily from the paper by David Capper et al. entitled [DNA methylation-based classification of central nervous system tumours](https://www.nature.com/articles/nature26000).

Download the data from GEO to local directory.

```{r data_dir, eval=TRUE}
# Download data
gse_acc <- "GSE97529"
data_dir <- "~/Downloads/GSE97529"
```

```{r download, eval=FALSE}
rgset <- yamat::get_gse(gse_acc, data_dir)
save(rgset, file = file.path(data_dir, "rgset.Rda"))
```

Normalization. I do not carry out batch effect correction or filtering out probes in this example. But, it is recommended to perform them in real analyses.

```{r norm, eval=FALSE}
# Load pre-calculated.
load(file.path(data_dir, "rgset.Rda"))

# Normalization
gmset <- yamat::normalize(rgset = rgset, norm_method = "swan")
```

Obtain the beta values.

```{r beta, eval=FALSE}
# Beta values
beta_vals <- minfi::getBeta(gmset, offset = 100)
save(beta_vals, file = file.path(data_dir, "beta_vals.Rda"))
```

Obtain and tidy the phenotype data. I add an column of tumor type abbreviations.

```{r tidy_pheno, eval=FALSE}
# Load pre-calculated.
load(file.path(data_dir, "beta_vals.Rda"))
load(file.path(data_dir, "rgset.Rda"))

# Phenotype
pheno_df <- minfi::pData(rgset)
pheno_df$tumor_type <- as.factor(pheno_df$`diagnosis:ch1`) %>%
  mapvalues(
    .,
    from = c("Ewing’s sarcoma", "Osteosarcoma", "Synovial sarcoma"),
    to = c("EWS", "OS", "SS")
  )
save(pheno_df, file = file.path(data_dir, "pheno_df.Rda"))
```

# Unsupervised analysis
## Most variably methylated loci

I use the most variably methylated loci in the unsupervised analysis. I choose 5000 most variable loci. You can choose a different number for your data through exploratory analysis, taking account of the number of samples.

```{r most_variable_loci, eval=TRUE}
load(file.path(data_dir, "beta_vals.Rda"))
load(file.path(data_dir, "pheno_df.Rda"))

top_n <- 5000
beta_vals_mv <-
  yamatClassifier::most_variable(beta_vals, top_n = top_n)
```

```{r figLociBetaSD, eval=TRUE, fig.cap="Distribution of standard deviations of beta values across loci. The vertical line is the threshold of 5000 most variable loci."}
yamatClassifier::density_plot_row_sd(beta_vals, top_n = top_n)
```

## Correlation between samples

I use the most variably methylated loci to calculate pairwise Pearson's correlation coefficients between samples, carry out hierarchical clustering, and visualize the result in heat map.

```{r reorder_samples, eval=TRUE}
beta_cor <- cor(beta_vals_mv)
dd <- as.dist((1 - beta_cor) / 2)
hc <- hclust(dd)
pheno_df <- pheno_df[hc$order, ]
beta_cor <- beta_cor[hc$order, hc$order]
beta_vals_mv <- beta_vals_mv[, hc$order]
beta_vals <- beta_vals[, hc$order]
```

```{r plot_corr, eval=TRUE}
# mat : is a matrix of data
# ... : further arguments to pass to the native R cor.test function
cor.mtest <- function(mat, ...) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p.mat<- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}
# matrix of the p-value of the correlation
p.mat <- cor.mtest(beta_vals_mv)


col <-
  colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
pal <- RColorBrewer::brewer.pal(3, "Dark2")

beta_cor2 <- beta_cor
data.frame(geo_accession = colnames(beta_cor2)) %>%
  dplyr::left_join(as.data.frame(pheno_df), by = "geo_accession") %>%
  dplyr::select(tumor_type) %>%
  unlist() %>%
  as.factor() -> tumor_type
rownames(beta_cor2) <- colnames(beta_cor2) <- tumor_type

corrplot(
  beta_cor2,
  method = "color",
  col = col(200),
  order = "original",
  tl.col = pal[tumor_type],
  tl.pos = "l",
  # Combine with significance
  p.mat = p.mat,
  sig.level = 0.01,
  insig = "pch",
  pch.cex = 0.5,
  pch.col = "grey70",
  # hide correlation coefficient on the principal diagonal
  diag = TRUE
)
```


