---
title: "Solid Tumor Example"
description: |
  A workflow to develop a classifier of a subset of solid tumors 
  from methylation array data.
author:
  - name: Mark Chen
    url: https://markgene.github.io/
    affiliation: SJCRH
    affiliation_url: https://www.stjude.org/
date: "`r Sys.Date()`"
citation_url: https://github.com/markgene/yamatClassifier
output: 
  radix::radix_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

```{r load_lib, message=FALSE, echo=FALSE, eval=TRUE}
library(yamat)
library(yamatClassifier)
library(ggplot2)
library(ggpubr)
library(corrplot)
library(RColorBrewer)
library(plyr)
library(dplyr)
library(RSpectra)
library(Rtsne)
```


# Data

I use the data from the paper by S. Peter Wu et al. entitled [DNA Methylation–Based Classifier for Accurate Molecular Diagnosis of Bone Sarcomas](http://ascopubs.org/doi/full/10.1200/PO.17.00031). It has 36 samples. However, I borrow the methodology manily from the paper by David Capper et al. entitled [DNA methylation-based classification of central nervous system tumours](https://www.nature.com/articles/nature26000).

Download the data from GEO to local directory.

```{r data_dir, eval=TRUE}
# Download data
gse_acc <- "GSE97529"
data_dir <- "~/Downloads/GSE97529"
```

```{r download, eval=FALSE}
rgset <- yamat::get_gse(gse_acc, data_dir)
save(rgset, file = file.path(data_dir, "rgset.Rda"))
```

Normalization. I do not carry out batch effect correction or filtering out probes in this example. But, it is recommended to perform them in real analyses.

```{r norm, eval=FALSE}
# Load pre-calculated.
load(file.path(data_dir, "rgset.Rda"))

# Normalization
gmset <- yamat::normalize(rgset = rgset, norm_method = "swan")
```

Obtain the beta values.

```{r beta, eval=FALSE}
# Beta values
beta_vals <- minfi::getBeta(gmset, offset = 100)
save(beta_vals, file = file.path(data_dir, "beta_vals.Rda"))
```

Obtain and tidy the phenotype data. I add an column of tumor type abbreviations.

```{r tidy_pheno, eval=FALSE}
# Load pre-calculated.
load(file.path(data_dir, "beta_vals.Rda"))
load(file.path(data_dir, "rgset.Rda"))

# Phenotype
pheno_df <- minfi::pData(rgset)
pheno_df$tumor_type <- as.factor(pheno_df$`diagnosis:ch1`) %>%
  mapvalues(
    .,
    from = c("Ewing’s sarcoma", "Osteosarcoma", "Synovial sarcoma"),
    to = c("EWS", "OS", "SS")
  )
save(pheno_df, file = file.path(data_dir, "pheno_df.Rda"))
```

# Unsupervised analysis
## Most variably methylated loci

I use the most variably methylated loci in the unsupervised analysis. I choose 5000 most variable loci. You can choose a different number for your data through exploratory analysis, taking account of the number of samples.

```{r most_variable_loci, eval=TRUE}
load(file.path(data_dir, "beta_vals.Rda"))
load(file.path(data_dir, "pheno_df.Rda"))

top_n <- 5000
beta_vals_mv <-
  yamatClassifier::most_variable(beta_vals, top_n = top_n)
```

```{r figLociBetaSD, eval=TRUE, fig.cap="Distribution of standard deviations of beta values across loci. The vertical line is the threshold of 5000 most variable loci."}
yamatClassifier::density_plot_row_sd(beta_vals, top_n = top_n)
```

## Correlation between samples

I use the most variably methylated loci to calculate pairwise Pearson's correlation coefficients between samples, carry out hierarchical clustering, and visualize the result in heat map.

```{r reorder_samples, eval=TRUE}
pheno_df <- pheno_df[match(colnames(beta_vals_mv), rownames(pheno_df)), ]
yamatClassifier::correlogram(beta_vals_mv, pheno = pheno_df$tumor_type)
```


## Principal component analysis (PCA)

Follow Capper's paper and the instructions on [Statistical Tools for High-Throughput Data Analysis](http://www.sthda.com/english/wiki/print.php?id=206), I carry out PCA in the following steps.

1. Center and scale.
2. Compute the correlation/covariance matrix.
3. Calculate the eigenvectors and eigenvalues.
4. Choose the PC number. I use Capper's method and fraction of
variance to calculate PC numbers and choose the bigger one
from the two methods.
5. Project the scaled input matrix onto the new basis.

I will transpose the matrix because it is more common to use columns as features and rows as samples in most R functions.

```{r pca, eval=TRUE}
t(beta_vals_mv) %>%
  yamatClassifier::pca(
    x = .,
    k = 50,
    seed = 1,
    threshold = 0.9
  ) -> pca_res
beta_vals_prj <- pca_res$projected
```

## tSNE

```{r tsne, eval=TRUE}
# Sets seed for reproducibility
set.seed(125) 
# Run tSNE
tsne_out_lst <-
  yamatClassifier::tsne(beta_vals_prj, perplexity = 5)

# Plot one tSNE result
pheno_df <- pheno_df[match(rownames(beta_vals_prj), rownames(pheno_df)), ]

plot_tsne(tsne_out_lst[[1]],
          colour = pheno_df$tumor_type,
          label = pheno_df$geo_accession)

# Plot multiple tSNE results
set.seed(456)
idx <- sample(x = seq(length(tsne_out_lst)), size = 5)
multiplot_tsne(tsne_out_lst[idx], pheno = pheno_df$tumor_type)

# Diagnose tSNE with coordinates plot
diagnose_tsne.coord(tsne_out_lst, pheno = pheno_df$tumor_type)

# Diagnose tSNE with correlation plot
diagnose_tsne.cor(tsne_out_lst, pheno = pheno_df$tumor_type)
```

